---
title: "Something Interesting Baseball"
author: "Jozi McKiernan, Sean O'Keeffe & Sam Woodman"
output: pdf_document
header-includes:
  - \usepackage{graphicx}
bibliography: report.bib
nocite: |
  @ggplot2, @shiny, @dplyr, @RCurl, @jsonlite, @shinydashboard, @RgoogleMaps, @xml, @tidyr, @mosaic, @stringr, @R
---

[//]: # Introduction Section-------------------------------------------------
# Introduction
Root, root, root for the home team - but what if the home team isn't who it used to be? Major League baseball players are traded every season by teams hunting for an edge. Imagine you could predict when your favorite players were going to change teams. Admit it - if you're a baseball fan, you've been puzzled by the logic of GMs as they barter prospects for big names and flip the roster upside down right before a deadline. 
What factors play a role in who gets traded? Obviously, there are some player-external factors - how the team is doing, managerial tendencies, and the state of the player's contract, to name a few - but what about factors related to the player's performance? Surely those play a role as well. And conveniently, there are a whole heap of statistics collected on the performance of every player in the major leagues. So let's check it out - can we predict whether a player will get traded this season based on their current performance? 

[//]: # Data Collection Section----------------------------------------------
# Data Collection
Unfortunately, the official MLB webpages and related fan pages were not created for data scientists. So step one is scraping the data from a number of pages and wrangling it into a usable format. To keep things reasonable, we decided to limit our search to records from the last five years only (2010-2015, inclusive). We found a list 
First of all, we need to collect the 
The data collection for this project was the most involved component.
We were able to find comprehensive statistics for any given player from @stats, however each page was for one player.
So we had to go find a list of all players for a given year.
We found this from @players.
Now that we had all the predicting attributes, we still needed to know if a player was traded.
All transactions for a given month for a given year was found through @trades.
However, there are instances of teams changing names and some entries from @trades, include minor league teams so we also needed a list of all the major league baseball teams and their naming history.
In this section, we will discuss some of the subtleties that made this a difficult process.

To keep our project organized, we split up the data collection into their own files. So we run all the data collection here and we will have code bits to describe some of the steps. Refer to the `final_project.R` file and all the other supplemental files.
```{r, message=FALSE, eval=FALSE}
setwd("~/Documents/GitHub/MATH154_FinalProject")
source("final_project.R", local=TRUE, echo=FALSE)
```

## List of all players

## Player statistics

## Teams
The list of team names were acquired manually, there are only 30 teams so it was very manageable.
We also got data on the location of every team.
Although a location is given in the team name, situations like the *Florida Marlins* is not an easy thing to locate on a map.
These locations were also acquired manually.
Team colors were also recorded manually.
Lastly, we also wanted latitude and longitude locations for each team.
This data came from google maps, using a package in R called `Rgooglemaps` @RgoogleMaps.
For each of the city locations, we look up the lat lon data with an API call to Google Maps.

The data looks like
```{r, eval=FALSE}
head(teams, n = 3)
```

## Trades
The trade data was a unique problem that we hadn't seen before.
When you try to scrape the HTML code for this website as we have done previously, no tables appear.
It turns out that the tables are generated on the client side through Java script so scraping the initial request doesn't give you anything. 
However the data itself had to be sent at some point, so after digging through some of the other files transfered, we found a JSON file that seemed to contain all the data we wanted.
So we query the URL for this file rather than the website.
```{r, eval=FALSE}
# This website loads the table data using Javascript. We go directly to the 
# JSON file which provides the data.
url <- paste('http://mlb.mlb.com/lookup/json/named.transaction_all.bam?start_date=',
             start_date,
             '&end_date=',
             end_date,
             '&sport_code=%27mlb%27',
             sep='')
json <- getURL(url)
tempData <- as.data.frame(fromJSON(json))
```
From here we trim down the data so that we only keep the data we're interested in.
```{r, eval=FALSE}
# Filter all the things we don't want. 
tempData <- tempData %>%
  # Only want trades
  filter(transaction_all.queryResults.row.type_cd=='TR') %>% 
  # Only want certain columns
  select(transaction_all.queryResults.row.team, 
         transaction_all.queryResults.row.from_team, 
         transaction_all.queryResults.row.player) %>% 
  # Remove any empty rows
  filter(transaction_all.queryResults.row.player != '') %>% 
  # Add a year attribute
  mutate(Year = year) %>% 
  # Add a key to match with stats data
  mutate(MatchKey = paste(transaction_all.queryResults.row.player, Year)) 
```

We do these steps for all 12 months during 2010 to 2015.
This ultimately gives us a table that is shown below, see .
```{r, eval=FALSE}
head(trades, n = 3)
```

[//]: # Classification Section-----------------------------------------------
# Classification
We used logistic regression to classify the data. 
To generate our logistic regression model, we used the glm() function on our training data. 
Then we used the model to predict whether or not players in the test data were traded. 

According to the coefficients of our model for pitchers, the contributions of the predictors rank as follows:
WHIP

BB9, H9, HR9, Year, W-L%

SHO, CG, L,R,  W,  GF, ERA, SV, ER, HBP, BK, SO9





[//]: # Shiny Web App Section------------------------------------------------
# Shiny Web App
The Shiny Web App has 3 tabs, Predict Trades, Trades Map, and Something.
Each one of these tabs will be described below.

## Predict Trades
We built a classifier using statistics of the baseball players to try and predict wether or not a player would get traded.
So in this tab you are able to input all the statistics that were used to build the classifier.
Then it will output a prediction of yes they will be traded or no they won't be.

## Trades Map
This is a graphic that lets you visualize the trades.
As you can see in Figure~\ref{fig:trades_map}, you can select any number of teams you want and on the map will be all the trades that happened during 2010 to 2015 to that team.
Below that is a data dump of those trades.
This doesn't tell us anything particularly useful about the data, it was just a fun visualization.

\begin{figure}
\centering
\includegraphics[width=0.6\textwidth]{trades_map.png}
\label{fig:trades_map}
\caption{Screenshot of Trades Map tab.}
\end{figure}


## Something
Maybe one variable dependence..?

\newpage
